Combining of local and global search in the parallel nested optimization scheme \thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2020-02-07 2020/PAVT_Nested_Full/pavt_paper.tex
Before proof-reading

The problem (\ref{ex_prob}) can be written in a recursive form 
\begin{equation}\label{outer_prob}
\varphi(\omega^*) = \min_\omega \varphi(\omega)
\end{equation}
\begin{equation}\label{inner_prob}
\varphi(\omega) = \min_c \Delta(\omega,c).
\end{equation}
The nested subproblem (\ref{inner_prob}) is a classical linear least squares problem, and its solution 
can be obtained by solving a system of linear algebraic equations regarding the unknown $c$, which 
can be done e.g. by Gaussian elimination. At the same time, the external problem (\ref{outer_prob}) 
will be, in general, a multiextremal one, and its solving will require considerable 
computational resources.

@@

%English begin
Suppose $n\geq 1$  iterations of the method have been carried out, in which trials were performed at 
$k=k(n)$ points $x^i,1\leq i \leq k$. Then points $x^{k+1},\dots,x^{k+p}$  of search trials of the next 
iteration are determined according to the following rules.

@@

Rule 3. For each interval ($x_{i-1},x_i), \; 2 \leq i \leq k,$ compute
the \textit{characteristics} $R(i)$ :
\begin{equation}\label{Rule_R}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r^2 \mu^2\Delta_i}-2\frac{z_i+z_{i-1}}{r \mu},
\end{equation}
where $\Delta_i=(x_i-x_{i-1})^{1/N}$. The value $r > 1$ is parameter of the algorithm. An 
appropriate selection
of $r$ allows to consider the product $r \mu$ as an estimate
of the Lipschitz constant $L$ of the objective function .

@@

Various modifications of this algorithm and the corresponding theory of
convergence are given in \cite{Barkalov2018,Sergeyev2013,Strongin2018,Strongin2000}.
%English end

@@

@@

The results of experiments performed on the series of the optimization problems
%, in which a multiextremal part and a unimodal one can be selected,
demonstrated the scheme of the 
parallel computations proposed in the present work to have a good potential for parallelization. 
In particular, when solving a series of 100 problems with total number of variables equal to 52 (from 
which, two first variables cause a global effect on the objective function and fifty affected this one 
locally), a speedup of 24.6 has been obtained when using  32 processes. 
In the case of solving  the problems with more complex local subproblems (with the number of the 
local variables $N=100$), the speedup increased up to 27.4 when using 32 processes.
The developed scheme of parallel computations can be applied in the identification of complex 
mathematical models, which are featured by a large number (tens and hundreds) of unknown 
parameters.

>>>>

The problem (\ref{ex_prob}) can be written in a recursive form 
\begin{equation}\label{outer_prob}
\varphi(\omega^*) = \min_\omega \varphi(\omega),
\end{equation}
\begin{equation}\label{inner_prob}
\varphi(\omega) = \min_c \Delta(\omega,c).
\end{equation}
The nested subproblem (\ref{inner_prob}) is a classical linear least squares problem, and its solution 
can be obtained by solving a system of linear algebraic equations regarding the unknown $c$, which 
can be done e.g. by Gaussian elimination. At the same time, the external problem (\ref{outer_prob}) 
will be, in general, a multiextremal one, and its solving will require considerable 
computational resources.

@@

Suppose $n\geq 1$  iterations of the method have been carried out, in which trials were performed at 
$k=k(n)$ points $x^i,1\leq i \leq k$. Then points $x^{k+1},\dots,x^{k+p}$  of search trials of the next 
iteration are determined according to the following rules.

@@

Rule 3. For each interval ($x_{i-1},x_i), \; 2 \leq i \leq k,$ compute
the \textit{characteristics} $R(i)$ :
\begin{equation}\label{Rule_R}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r^2 \mu^2\Delta_i}-2\frac{z_i+z_{i-1}}{r \mu},
\end{equation}
where $\Delta_i=(x_i-x_{i-1})^{1/N}$. The value $r > 1$ is parameter of the algorithm. An 
appropriate selection
of $r$ allows to consider the product $r \mu$ as an estimate
of the Lipschitz constant $L$ of the objective function.

@@

Various modifications of this algorithm and the corresponding theory of
convergence are given in \cite{Barkalov2018,Sergeyev2013,Strongin2018,Strongin2000}.
%Замечание по поводу теоретических свойств и сравнения алгоритмов
The results of the experimental comparison  (see, for example, \cite{Sovrasov2019}) show that global search algorithm is superior to many known methods of similar purpose, including both deterministic and heuristic  algorithms.

@@

% В последней серии экспериментов было решено 10 задач, в которых в качестве многоэкстремальной части использовались трехмерные задачи из класса GKLS. Данный класс задач часто используется для тестирования методов глобальной оптимизации [ссылки]. Размерность локальных подзадач была равна N = 100. Задачи решались на 1, 8, 16 и 32 узлах кластера, на которых использовалось от 4 до 128 MPI-процессов. Параметры запуска такие же, как в предыдущих сериях экспериментов. Результаты решения представлены в таблице, ускорение посчитано относительно запуска на одном узле на 4 mpi процесса.

In the last series of experiments, 10 problems were solved, in which three-dimensional functions from GKLS class \cite{Gaviano2003} were used as a multiextremal part. 
This class of functions is often used for investigating global optimization algorithms \cite{Sergeyev2015,Paulavicius2014,Barkalov2018}.
The dimensionality of the local subproblems was equal $N=100$.
The problems were solved at 1, 8, 16 and 32 cluster nodes, on which from 4 to 128 MPI processes were used.
The running parameters were the same as in the previous series of experiments.
The results of solving are presented in Table \ref{tab3}, where speedup is calculated relative to running on a single node using 4 MPI processes.

%In the last series of experiments, 10 problems of the kind (\ref{test_problem}) were solved, in which the dimensionality of the local subproblems was equal $N=100$. The running parameters were the same as in the previous series of experiments. The results of solving are presented in Table \ref{tab3}.

\begin{table}
\centering
\caption{Results of solving a series of problems of dimensionality $N=103$}\label{tab3}
\begin{tabular}{cccc}
\hline\noalign{\smallskip}
 $\;\;\;p\;\;\;$  &  $\;\;\;K_{av}\;\;\;$ &  $\;\;\;T_{av}\;\;\;$ & $\;\;\;S\;\;\;$ \\
\noalign{\smallskip}\hline\noalign{\smallskip}
 4  & 730.8  & 63.4 & ---  \\
 32 & 87.4   & 7.37 & 8.6 \\
 64 & 37.0   & 3.77 & 16.8 \\
 128 & 16.5   & 1.78 & 35.7 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

The results of experiments performed on the series of the optimization problems
%, in which a multiextremal part and a unimodal one can be selected,
demonstrated the scheme of the 
parallel computations proposed in the present work to have a good potential for parallelization. 
In particular, when solving a series of 100 problems with total number of variables equal to 52 (from 
which, two first variables cause a global effect on the objective function and fifty affected this one 
locally), a speedup of 24.6 has been obtained when using  32 processes. 
In the case of solving  the problems with more complex local subproblems (with the number of the 
local variables $N=100$), the speedup increased up to 35 when using 128 processes. In the last case the speedup was calculated with respect to the use of 4 processes on one node.
The developed scheme of parallel computations can be applied in the identification of complex 
mathematical models, which are featured by a large number (tens and hundreds) of unknown 
parameters.
